Continuation Passing Style for Effect Handlers

### 2020-02-02 docs/article.tex
Update article

\title{Continuation Passing Style for Effect Handlers}
\author{Mateusz Urbańczyk}
\date{7 January 2020}

@@

    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
    $V \oplus W$ and $V \approx W$ are respectively binary and relational
    operators, where we support basic arithmetic and comparison operations.
    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching statement.
    The other constructs are just as in Links, with slight syntax modifications.
    Actual programs in Freak can be found in \fullref{sec:examples}

@@

    \begin{flushleft}
    $\mathcal{E}$ ::= $ \mathcal{E} \oplus W \; | \; V \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$ \\
    \end{flushleft}

@@

\section{Usage guide}
    % TODO: not sure if correct place for that

    All commands are available within \verb src  directory.

@@

    \begin{itemize}
        \item Install dependencies: \verb!make install!
        \item Compile: \verb!make build!
        \item Link to PATH: \verb!sudo make link!
        \item Remove artiacts: \verb!make clean!
    \end{itemize}

    After compiling and linking program to PATH, one may evaluate program as
    follows: \verb!freak programs/choicesList.fk!. The actual code is described in \fullref{sec:choice-example}

@@

\section{Implementation}

    The Freak implementation is available \href{https://github.com/Tomatosoup97/freak}{\underline{here}}.
    While the paper provided a good overview of the language and the translation,
    the lower-level details were omitted. Therefore, the actual CPS implementation
    is based on A. Appel \cite{appel-continuations}, which provides a translation
    for a simplified ML calculus. \\

    Abstract Syntax Tree structure is defined just as syntax is. The core data
    structure, into which the source program is transformed, is defined as follows:

    \begin{verbatim}
data ContComp
    = CPSApp CValue [CValue]
    | CPSFix Var [Var] ContComp ContComp
    | CPSBinOp BinaryOp CValue CValue Var ContComp
    | CPSValue CValue
    | CPSLet Var CValue ContComp
    | CPSSplit Label Var Var CValue ContComp
    | CPSCase CValue Label Var ContComp Var ContComp
    | CPSIf CValue ContComp ContComp
    | CPSAbsurd CValue
    \end{verbatim}

    The translation is based on the curried first-order translation. For
    improving the performance of the evaluation, uncurried higher-order translation
    should be adopted, so that administrative redexes are contracted and
    proper tail-recursion is obtained. The final answer is represented as a \verb!DValue!,
    where the meaning of the coproduct is as one would expect:

    \begin{verbatim}
type Label = String

data DValue
    = DNum Integer
    | DLambda Env FuncRecord
    | DUnit
    | DPair DValue DValue
    | DLabel Label
    \end{verbatim}

@@

    \begin{verbatim}
    AST.hs          - AST data structures
    CPS.hs          - CPS translation
    Eval.hs         - Evaluation of the target calculus
    Freak.hs        - API for the language
    Main.hs         - Main module running evaluator on given filename
    Parser.hs       - Parser and lexer
    Tests.hs        - Tests module
    Typechecker.hs  - -- suspended --
    Types.hs        - Common types definition
    programs/       - Exemplary programs used in tests
    \end{verbatim}

@@

    These implications are straightforward when looking from a categorical
    standpoint and models of algebraic theories, nevertheless are very exciting
    for programming use cases. The current Freak implementation does not yet support I/O.

>>>>

\title{Continuation Passing Style for Effect Handlers}
\author{Mateusz Urbańczyk}
\date{31 January 2020}

@@

    The syntax for the calculus is shown below. $nat \; n$ represents an integer $n$,
    $V \oplus W$ and $V \approx W$ are respectively binary and relational
    operators, where we support basic arithmetic and comparison operations.
    \textbf{if} $V$ \textbf{then} $M$ \textbf{else} $N$ is a standard branching statement.
    The other constructs are just as in Links, with slight syntax modifications.
    Actual programs in Freak can be found in section \ref{sec:examples}.

@@

    \begin{flushleft}
    $\mathcal{E}$ ::= $ \mathcal{E} \oplus W \; | \; nat \; n \oplus \mathcal{E} \; |$ \textbf{if} $\mathcal{E}$ \textbf{then} $M$ \textbf{else} $N$ \\
    \end{flushleft}

@@

\section{Usage guide}

    As of this day, two implementations are available, one based on the curried
    translation and Appel \cite{appel-continuations}, and the second one based
    directly on the uncurried translation with continuations as explicit stacks
    from paper. More details can be found in section \ref{sec:implementation}.
    All commands are available within \verb src  directory.

@@

    \begin{itemize}
        \item Install dependencies: \verb!make install!
        \item Select implementation: \verb!make link-lists! (default) vs \verb!make link-appel!
        \item Compile: \verb!make build!
        \item Link to PATH: \verb!sudo make link!
        \item Remove artifacts: \verb!make clean!
    \end{itemize}

    After compiling and linking program to PATH, one may evaluate program as
    follows: \verb!freak programs/choicesList.fk!. The actual code is described in section \ref{sec:choice-example}

@@

\section{Implementation}
\label{sec:implementation}

    The Freak implementation is available \href{https://github.com/Tomatosoup97/freak}{\underline{here}},
    written purely in Haskell. While the paper provided a good overview of the
    language and the translation, the lower-level details were omitted. That
    being said, two inherently different takes at the implementations were made.
    The first one is based on curried translation and A. Appel \cite{appel-continuations}
    book, and the second one directly on the uncurried translation to target
    calculus with continuations represented as explicit stacks from the paper.
    We start by presenting core data structrues, and afterwards move to actual
    translation details.

    \subsection{Abstract Syntax Trees}
    \label{sec:ast}

    The language's AST is defined without surprises, just as syntax is:

    \begin{verbatim}
data Value
    = VVar Var
    | VNum Integer
    | VLambda Var ValueType Comp
    | VFix Var Var Comp
    | VUnit
    | VPair Value Value
    | VRecordRow (RecordRow Value)
    | VExtendRow Label Value Value
    | VVariantRow (VariantRow Value)
    | VBinOp BinaryOp Value Value

data Comp
    = EVal Value
    | ELet Var Comp Comp
    | EApp Value Value
    | ESplit Label Var Var Value Comp
    | ECase Value Label Var Comp Var Comp
    | EReturn Value
    | EAbsurd Value
    | EIf Value Comp Comp
    | EDo Label Value
    | EHandle Comp Handler
    \end{verbatim}
    Similarly for the target calculus data structure. However, as one may notice,
    for convenience the \textbf{let} translation is homomorphic, as opposed to be
    to lambda abstracted with immediate application:

    \begin{verbatim}
data UValue
    = UVar Var
    | UNum Integer
    | UBool Bool
    | ULambda Var UComp
    | UUnit
    | UPair UValue UValue
    | ULabel Label
    | URec Var Var UComp
    | UBinOp BinaryOp UValue UValue

data UComp
    = UVal UValue
    | UApp UComp UComp
    | USplit Label Var Var UValue UComp
    | UCase UValue Label UComp Var UComp
    | UIf UValue UComp UComp
    | ULet Var UComp UComp
    | UAbsurd UValue

    \end{verbatim}
    The final answer, common to both evaluations, is represented as a \verb!DValue!,
    where the meaning of the coproduct is as one would expect:

    \begin{verbatim}
type Label = String
type FuncRecord = [DValue] -> Either Error DValue
data DValue
    = DNum Integer
    | DLambda FuncRecord
    | DUnit
    | DPair DValue DValue
    | DLabel Label
    \end{verbatim}

    \subsection{Curried translation}

    The first take was heavily inspired by A. Appel's Compiling with Continuations
    \cite{appel-continuations}, which provides a translation for a simplified ML
    calculus. The calculus was extended and translation adopted to handle algebraic
    effects and their handlers. The translation is based on the curried first-order
    translation. That being said, the source code diverged a lot from the paper on
    which it was based, leading to a different transformation for which
    the correctness and cohesion with operational semantics should be proved
    separately. Indeed, while the interpreter worked well on the use cases
    defined in tests, the evaluation had part which was not tail-recursive. What's
    more, nested handlers were not supported, and the implementation was found to be
    more tricky than it should, as it was not obvious on how to adopt the technique
    proposed in the paper. \\

    In terms of improving the performance of the evaluation, uncurried higher-order
    translation should be adopted, so that administrative redexes are contracted
    and proper tail-recursion is obtained. The core data structure, into which the
    source program is transformed, is defined as follows:

    \begin{verbatim}
data ContComp
    = CPSApp CValue [CValue]
    | CPSResume CValue ContComp
    | CPSFix Var [Var] ContComp ContComp
    | CPSBinOp BinaryOp CValue CValue Var ContComp
    | CPSValue CValue
    | CPSLet Var CValue ContComp
    | CPSSplit Label Var Var CValue ContComp
    | CPSCase CValue Label Var ContComp Var ContComp
    | CPSIf CValue ContComp ContComp
    | CPSAbsurd CValue
    \end{verbatim}
    Each term at the end has a coinductive reference to itself, which represents
    the rest of the computation that needs to be done. For more clarification,
    one may take a look that the book cited above. The source code for curried
    translation and evaulation can be found respetively in \verb!CPSAppel.hs! and
    \verb!EvalCPS.hs!.

    \subsection{Uncurried translation}

    Having in mind the drawbacks mentioned above, alternative translation was
    written, that coincides with the translation from the paper. Namely, with
    the uncurried translation to target calculus with continuations represented
    as explicit stacks. The target calculus was described in section \ref{sec:ast},
    for which the evaluation is straightforward. The continuations are represented
    as \verb!Cont!, with syntactic distinction between pure and effectful computations,
    which occupy alternating positions in the stack. Explicit distinction gave
    more control in the source code.

    \begin{verbatim}
type CPSMonad a = ExceptT Error (State Int) a

type ContF = UValue -> [Cont] -> CPSMonad UComp

data Cont = Pure ContF
          | Eff ContF
    \end{verbatim}
    Where \verb!CPSMonad! is a monad transformer over \verb!Either! and \verb!State!.
    \verb!State! was required to generate labels for fresh variables that
    came from the translation. The core code is split into five functions

    \begin{verbatim}
cps     :: Comp    -> [Cont] -> CPSMonad UComp
cpsVal  :: Value   -> [Cont] -> CPSMonad UValue
cpsHRet :: Handler -> Cont
cpsHOps :: Handler -> Cont
forward :: Label   -> UValue -> UValue -> [Cont] -> CPSMonad UComp
    \end{verbatim}
    Where the first two are implementing cps for computations and values.
    \verb!cpsHRet! and \verb!cpsHOps! are yielding pure and effectful computation,
    based on a given handler. The last one is responsible for forwarding the
    computation to the outer handler.\\

    This results in an implementation that finally supports nested handlers,
    as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
    Unfortunatelly, following closely translation from the paper resulted in
    behaviour in which invoked resumption forgets its pure continuation. This means,
    that the following code, evaluating correctly on the Appel-based translation,
    returns 0 rather than 1:

    \begin{verbatim}
handle do Drop () with { Drop p r -> let t <- r 0 in return 1}
    \end{verbatim}
    Nevertheless, working out this issue seems less demanding than coping with
    discrepancies created in the first translation. The source code for uncurried
    translation and evaulation can be found respetively in \verb!CPSLists.hs! and
    \verb!EvalTarget.hs!.

@@

    \begin{verbatim}
    AST.hs          - AST data structures
    CommonCPS.hs    - Common functions for CPS translation
    CommonEval.hs   - Common functions for evaluation
    CPSAppel.hs     - Appel based CPS translation
    CPSLists.hs     - Uncurried CPS translation
    EvalCPS.hs      - Evaluation of the Appel's CPS structure
    EvalTarget.hs   - Evaluation of the target calculus
    Freak.hs        - API for the language
    Main.hs         - Main module running evaluator on given filename
    Parser.hs       - Parser and lexer
    TargetAST.hs    - AST for the target calculus
    Tests.hs        - Tests module
    Types.hs        - Common types definition
    programs/       - Exemplary programs used in tests
    \end{verbatim}

@@

    These implications are straightforward when looking from a categorical
    standpoint, where effects are viewed as free models of algebraic theories
    \cite{adequacy}, and handlers are homomorhpisms preserving the model structure
    \cite{handlers}. Nevertheless, the results are very exciting for programming
    use cases. The current Freak implementation does not support I/O.
